---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r chunk1.loading, echo=TRUE, fig.path='figure/'}
library(dplyr)
library(ggplot2)

dataset <- read.csv(unz("activity.zip", "activity.csv"))
dataset$date <- as.Date(dataset$date, "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
Create function to show plot with mean and median, because we need to repeat it later
```{r chunk2.mean.per.day, echo=TRUE, result='asis', fig.path='figure/'}
showMeanMedianByDate <- function (showme) {
  # Group data by date and summarize it by taking sum of steps
  showme.bydate = showme %>% 
    filter(!is.na(steps)) %>%
    group_by(date) %>% 
    summarize(steps = sum(steps))
  
  print(ggplot(showme.bydate, aes(steps)) +
    labs(title="Histogram of the total number of steps taken each day") +
    labs(x="Steps", y="Days") +
    geom_histogram(bins=5, col="white"))
  
  cat(paste(
    "Mean and median of the total number of steps taken per day:",
    mean(showme.bydate$steps), 
    "and", 
    median(showme.bydate$steps)))
}

showMeanMedianByDate(dataset)
```


## What is the average daily activity pattern?
```{r chunk3.daily.activity, echo=TRUE, result='asis', fig.path='figure/'}
days.count <- length(unique(dataset$date))

dataset.bytime <- dataset %>%
  # Group data by interval and summarize it by using mean of steps
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))

with(dataset.bytime, plot(
  interval, 
  steps, 
  type="l",
  main="Average number of steps taken of 5-minute interval",
  xlab="Interval",
  ylab="Average steps"))

cat(paste(
  "5-minute interval, on average across all the days in the dataset,",
  "contains the maximum number of steps:",
  dataset.bytime[which.max(dataset.bytime$steps),]$interval))
```


## Imputing missing values
```{r chunk4.imputing, echo=TRUE, fig.path='figure/'}
print(paste(
  "Total number of missing values in the dataset:",
  sum(!complete.cases(dataset))
))

# NA values are only in dataset steps column
# Fill all missing values using mean for that 5-minute interval
# dataset.bytime contains averaged (by interval) number of steps
# So we need to merge these two tables together (by interval column)
# And fix steps column by replacing NA values to corresponding average
dataset.fixed <- merge(dataset, dataset.bytime, by="interval") %>%
  rename(steps=steps.x, steps.mean=steps.y) %>%
  mutate(steps=ifelse(is.na(steps),steps.mean,steps)) %>%
  arrange(date, interval)

# Make histogram of number of steps and report mean and median
showMeanMedianByDate(dataset.fixed)
```


Because we fixed NA values of steps by using average number of steps in the particular interval,
our mean value of number of steps is unchanged. In original dataset each number of steps is integer value, and median value for original dataset is integer too. But, in fixed dataset, we now have a lot of floating point values, so our median is now equal to mean.


## Are there differences in activity patterns between weekdays and weekends?
```{r chunk5.weekends.activity, echo=TRUE, fig.path='figure/'}
dataset.fixed$week <- weekdays(dataset.fixed$date) %in% c("Saturday", "Sunday")
dataset.fixed$week <- factor(dataset.fixed$week, labels=c("weekday","weekend"))

# Group data set by interval and week and summarize it using mean of steps
dataset.week <- dataset.fixed %>%
  group_by(interval, week) %>%
  summarize(steps = mean(steps))

# Because we grouped data by week column (weekend or weekday) we will get 2 plots
# Draw each plot as line by using interval and steps values
print(ggplot(dataset.week, aes(interval, steps)) + 
      geom_line() + 
      xlab("Interval") + ylab("Number of steps") +
      facet_wrap(~week, ncol=1))
```